Dynamics of Colombo's Top: Generating Exoplanet Obliquities from Planet-Disk Interactions

### 2020-08-13 writeups/response.txt

> 2) This is primarily a matter of preference, but I suggest considering the usage of the polar depiction of the phase space, as opposed to using \phi and \cos\theta as the x and y axes. The polar view can assist in seeing where and when the separatrix emerges. In addition, the calculations of phase space areas tend to be more intuitive, and it could also help the reader better interpret the scenarios in Figures 8-11.

Yes, there are definitely many advantages to the polar depiction of phase space. However, we tried in this paper to highlight the detailed behavior near separatrix crossing, and it seems much easier to zoom in to the vicinity of the separatrix in (\phi, \cos\theta) coordinates. There is precedent using these coordinates for this reason, e.g. https://doi.org/10.1093/mnras/stw3018. So we choose to retain the current coordinate system.

> 3) There is an interesting difference between separatrix crossings that change the obliquity with an impulsive and large jump (e.g. III->I, III->II->I) versus those that do not affect the obliquity evolution in a significant way (e.g. II->I, II->III). Is this distinction fundamental enough to be emphasized in the text?

This largely arises due to the choices of parameters used for Figs 7-8; a larger \theta_{sd, i} would show that II -> I,III crossings also impulsively change the obliquity behavior. As such, we choose not to remark on this in the paper.

@@

> 4) The whole set-up assumes a single planet in the disk. Can the authors comment (briefly) on what changes when multiple planets are present? Similarly, are there any interesting dependencies on the timing at which the resonance encounter occurs? (E.g. closer planets encounter the resonance earlier than more distant ones.)

We have added a brief discussion on multiple planet configurations, focusing on aspects for which our analysis differs from that of Millholland & Batygin 2018.

@@

Smaller comments:
> - State references for equations 3 and 4.

Added.

> - In equation 3 and 4, it is more customary in previous literature to label k_2 = 3 k_{qp}, where k_2 is the Love number. In addition, conventional notation for k_p is \lambda or C.

Our notation follows the notation used in our references, though we have added equivalences to the common choices of notation.

@@

> - Figures 6 and 7 look very similar, and it does not seem to add much to have both of them. I would suggest either removing one or making Figure 7 an even larger angle than I=20 deg.

Removed Figure 6, leaving only I = 20.

> - In section 3.2.2, whenever the probability is between 0 and 1, how does the phase factor into determining which track the system ends up following? Is the phase the only factor that influences which side the trajectory will come out on?

This is a great question. Strictly speaking, in the pure adiabatic limit, the phase cannot affect the evolution of the system. Instead, the result from Henrard 1982 (Eqs. 25) shows that the true outcome depends on the value of the Hamiltonian modulo a small quantity ~epsilon. When epsilon is sufficiently small, this is effectively probabilistic, rather than deterministic. We have added a footnote summarizing this relavant result.

@@

> - There is an extra partial derivative at the top of page 8 shortly before equation 30.

Removed.

> - Whether the area A_I is positive or negative depends on the direction of the trajectory in the phase space. It would be helpful to clarify this in a couple areas in Section 3.2.3. It is understandable upon examination of the figures, but at first look at the text, it is confusing why area A_I is sometimes denoted as negative (e.g. just before equation 26) and other times positive (e.g. just before equation 28).

We have added such a clarification. We believe this confusion is further remedied by the change in notation to differentiate between unsigned areas (a la Ward & Hamilton) and signed phase space areas, c.f. 6th small comment.

@@

> - On Figures 8-11 and 14, it might be helpful to maintain a uniform y-axis range, since the evolution of the phase space and separatrix crossings would be clearer.

We believe a very narrow y-axis is necessary to show the evolution of the trajectory relative to the changing separatrix and how separatrix crossing is executed. However, we note the reviewer's concern that the plots do not convey the phase space area evolution very well, so we choose to zoom the plots out slightly, to better convey the message that the shaded areas extend to the poles.

> - In Figure 13, why does the vertical banding not extend all the way to the green III->I track? Is it because the libration cycles deviate more strongly from circular in this regime?

This is a great observation. Our calculations only assume the libration cycles are circular initially; at later times, we use conservation of phase space area to compare the phase space area at any time to the analytical A_{I,II,III}s. Instead, this is because the separatrix is changing so much during the separatrix-crossing orbit, approximating eta_star as a constant during this orbit is invalid. We have made note of it in the manuscript.

### 2020-08-13 writeups/Su_disk_disp.tex
typos
finished paper response

% rm -f plots_diskdisp/*.png; for i in ../initial/99_misc/2_nonad_rot.png ../initial/99_misc/2_3vec.png ../initial/99_misc/2_cs_locs.png ../initial/0_eta/1contours_flip.png ../initial/99_misc/1_areas.png ../initial/2_toy2/3_ensemble_05_35.png ../initial/2_toy2/3_ensemble_20_35.png ../initial/2_toy2/3testo21.png ../initial/2_toy2/3testo21_subplots.png ../initial/2_toy2/3testo23.png ../initial/2_toy2/3testo23_subplots.png ../initial/2_toy2/3testo31.png ../initial/2_toy2/3testo31_subplots.png ../initial/2_toy2/3testo321.png ../initial/2_toy2/3testo321_subplots.png ../initial/2_toy2/3_ensemble_05_25.png ../initial/2_toy2/3_ensemble_05_15.png ../initial/2_toy2/3testo_nonad.png ../initial/2_toy2/3testo_nonad_subplots.png ../initial/2_toy2/3_ensemble_05_05.png ../initial/2_toy2/3scan.png ../initial/2_toy2/3scan_20.png ../initial/99_misc/2_lambdas.png; do cp $i plots_diskdisp; done
    \documentclass[twocolumn,twocolappendix]{aastex63}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
    }
    \usepackage[T1]{fontenc}

@@

In this paper, we consider a wide range of initial planetary obliquities in the
Millholland-Batygin dissipating disk scenario, and examine how the obliquity
evolves toward the ``final'' value as the exterior disk dissipates. We provide
an analytical framework for understanding the final planetary obliquity for
arbitrary initial spin-disk misalignment angles. We also consider various
dissipation timescales, and examine both ``adiabatic'' (slow disk dissipation)
and ``non-adiabatic'' evolution. \textcolor{Corr}{Our chief results include
that: (i) careful accounting of phase space area across separatrix
crossings is able to explain the rich dynamics of adiabatic evolution, and (ii)
the concept of ``partial adiabatic resonance advection'' fully explains dynamics
in the non-adiabatic limit.} We calibrate our analytical results with numerical
calculations.

@@

We define the dimensionless time $\tau$ as
\begin{equation}
    \tau \equiv \alpha t,
\end{equation}
and the frequency ratio $\eta$
\begin{align}
    \eta \equiv{}& -\frac{g}{\alpha}\nonumber\\
        ={}& 2.08 \p{\frac{k_{\rm p}}{k_{\rm qp}}}
            \p{\frac{\rho_{\rm p}}{\mathrm{g/cm}^3}}
            \p{\frac{M_{\rm d}}{0.01 M_{\odot}}}
            \p{\frac{a_{\rm p}}{5\;\mathrm{AU}}}^{9/2}\nonumber\\
        &\times
            \p{\frac{r_{\rm d}}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{M_{\odot}}}^{-3/2}
            \p{\frac{P_p}{10\;\mathrm{hrs}}}
            \cos I ,\label{eq:eta}
\end{align}
% g = 1 / (51249 yr) 1 / (3 / 4 * (0.01) / (30 AU)^3 * sqrt(G * (solar mass) * (5AU)^3))
% alpha = 1/(106700 yr) 1 / (3/2 * (1 solar mass / (5AU)^3) / (4*pi / 3 * 1g/cm^3) * (2pi / (10 hours)))
% ((3 * 0.01 / 4) * (1 / 30AU)^3 * (10 hours * sqrt(G * (solar mass)) / (2pi)) / (1.5 * (1 solar mass / (5AU)^(9/2)) / (4 * pi / 3 * (1g/cm^3))))
where $\rho_p = 3M_p/(4\pi R_p^3)$ and $P_p = 2\pi/\Omega_p$ is the planet's
rotation period. . In Eq.~\eqref{eq:eta}, we have introduced the fiducial values
of variable parameters for the application considered in this paper.
Eq.~\eqref{eq:dsdt_rot} then becomes
\begin{equation}
    \p{\rd{\uv{s}}{\tau}}_{\rm rot} = \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        - \eta\p{\uv{s} \times \uv{l}_{\rm d}}. \label{eq:dsdt_base}
\end{equation}

@@

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Trajectories in zone II librate about CS2 while those in
zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] for $I = 5^\circ$, where warmer colors denote more
    positive values. The black solid line is the separatrix, which only exists
    for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III), divided by
    the separatrix, are labeled. The Cassini states are denoted by filled
    circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The interior
    of the separatrix, shaded in grey, is formally only defined for $\eta <
    \eta_{\rm c}$, but we may identify the points in phase space that flow into
    zone II when evolved forward in time (decreasing $\eta$ adiabatically); this
    is the shaded region in panel (a), bounded by the black dotted
    line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically
\citep{henrard1987,ward2004I}. \textcolor{Corr}{As given in
\citet{ward2004I}, if we define}
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},
    \end{align}
\end{subequations}
\textcolor{Corr}{then} the areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        \mathcal{A}_{\rm I} &= 2\pi\p{1 - z_0} - \frac{\mathcal{A}_{\rm II}}{2},
            \label{eq:A1}\\
        \mathcal{A}_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan
            \frac{1}{\chi}, \label{eq:A2}\\
        \mathcal{A}_{\rm III} &= 2\pi\p{1 + z_0} - \frac{\mathcal{A}_{\rm
            II}}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$\mathcal{A}_{\rm j}\p{\eta > \eta_{\rm c}} = \mathcal{A}_{\rm j}(\eta_{\rm
c})$. The boundary between these extended zones is denoted by the dashed black
line in panel (a) of Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/1_areas.png}
    \caption{Plot of fractional areas of each of the zones
    $\mathcal{A}_{\rm i}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward} for
    $I = 5^\circ$. The colored dotted lines correspond to small $\eta$
    approximations used in Appendix~\ref{s:ad_approx}. The colored dashed lines
    for $\eta > \eta_{\rm c}$ are the effective values of $\mathcal{A}_{\rm II},
    \mathcal{A}_{\rm III}$ for $\eta > \eta_{\rm c}$, denoting the points that
    would flow into either area under adiabatic decrease of $\eta$ from $\eta >
    \eta_{\rm c}$ (see the text). The vertical black dashed lines correspond to
    $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}] and the values of $\eta$ for
    which $\mathcal{A}_{\rm II}$ is maximized ($\eta_{\rm \max, II}$) and for
    which $\mathcal{A}_{\rm III}$ is minimized ($\eta_{\rm \min, III}$,
    Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_05_35.png}
    \caption{Top: The final spin obliquity $\theta_{\rm f}$ as a function of the
    initial spin-disk misalignment angle $\theta_{\rm sd, i}$ for systems
    evolving from initial $\eta_{\rm i} \gg 1$ to $\eta_{\rm f} \ll 1$, for $I =
    5^\circ$. The blue dots are results of numerical calculations
    (Section~\ref{ss:ad_ensemble}), and the colored tracks are semi-analytical
    results (Section~\ref{ss:zone_transitions}). Bottom: The probabilities of
    different outcomes. Where a particular $\theta_{\rm sd, i}$ corresponds to
    multiple tracks, the system evolves probabilistically. The track that a
    particular system evolves along in a numerical simulation can be measured by
    examining its final obliquity. The dots represent the inferred probabilities
    from measured final obliquities in our simulations, while the colored tracks
    denote the semi-analytic probability of the system evolving along each
    track. There are five regimes of $\theta_{\rm sd, i}$ values for which
    different tracks are accessible. In both plots, the vertical dashed black
    lines denote semi-analytical calculations of the boundaries of these regimes
    (see Section~\ref{ss:zone_transitions}), while the black dotted lines
    represent analytical approximations valid in the small-$\theta_{\rm sd, i}$
    limit (see Appendix~\ref{s:ad_approx}).}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_20_35.png}
    \caption{Same as the top panel of Fig.~\ref{fig:ad_ensemble} but for $I
    = 20^\circ$ and with fewer annotations.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has two advantages: (i) it is continuous across
transitions from circulating to librating that cross the North pole ($\cos
\theta = 1$), and (ii) the areas of the three zones are equal in absolute value
to the expressions given in Eqs.~\eqref{se:area_ward}. The path over which the
integral is taken is either a libration or circulation cycle. When $\eta_{\rm i}
\gg 1$, trajectories librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm
sd}$, meaning they enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is studied by \citet{henrard1982} and
\citet{henrard1987}. Their results may be summarized as follows: if zone $i$
is shrinking while adjacent zones $j, k$ are expanding such that the sum of
their areas is constant, the probabilities of transition from zone $i$ to zones
$j$ and $k$ are given by
\begin{subequations}\label{eq:henrard_hop}
    \begin{align}
        \Pr\p{i \to j} = -\frac{\pdil{\mathcal{A}_{\rm
            j}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}, \\
        \Pr\p{i \to k} = -\frac{\pdil{\mathcal{A}_{\rm
            k}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}.
    \end{align}
\end{subequations}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$.
Eqs.~\eqref{eq:henrard_hop} can be used in conjunction with
Eqs.~\eqref{se:area_ward} to understand for what initial conditions each track
can be observed and with what probabilities.

@@

As $\epsilon$ increases (see Fig.~\ref{fig:3_ensemble_05_25}), nonadiabaticity
manifests as a larger scatter of final obliquities near the tracks predicted
from adiabatic evolution. This scatter first sets in for trajectories starting
in zone III, as these trajectories encounter the separatrix at larger $\eta$
compared to those originating in zone II\@. This means the obliquity of CS2
$\theta_2$ is smaller for these trajectories, and the adiabaticity criterion is
stricter [see Eq.~\eqref{eq:ad_constr}]. Physically, approaching the
adiabaticity criterion corresponds to the separatrix crossing process becoming
sensitive to the \emph{phase} of the libration/circulation cycle at the
crossing: if the trajectory crosses the separatrix when the obliquity is at its
maximum, the final obliquity will also be relatively larger.

As $\epsilon$ increases further (see Fig.~\ref{fig:3_ensemble_05_15}) but still
marginally satisfies the weak adiabaticity criterion [Eq.~\eqref{eq:ad_constr}],
the scatter in $\theta_{\rm f}$ continues to widen. The horizontal banded
structure of the final obliquities is a consequence of even stronger phase
sensitivity during separatrix crossing: trajectories cross the separatrix at
similar phases evolve to similar final obliquities that only depend weakly on
on $\theta_{\rm sd, i}$.

>>>>

% rm -f plots_diskdisp/*.png; for i in ../initial/99_misc/2_nonad_rot_cropped.png ../initial/99_misc/2_nonad_rot.png ../initial/99_misc/2_3vec.png ../initial/99_misc/2_cs_locs.png ../initial/0_eta/1contours_flip.png ../initial/99_misc/1_areas.png ../initial/2_toy2/3_ensemble_05_35.png ../initial/2_toy2/3_ensemble_20_35.png ../initial/2_toy2/3testo21.png ../initial/2_toy2/3testo21_subplots.png ../initial/2_toy2/3testo23.png ../initial/2_toy2/3testo23_subplots.png ../initial/2_toy2/3testo31.png ../initial/2_toy2/3testo31_subplots.png ../initial/2_toy2/3testo321.png ../initial/2_toy2/3testo321_subplots.png ../initial/2_toy2/3_ensemble_05_25.png ../initial/2_toy2/3_ensemble_05_15.png ../initial/2_toy2/3testo_nonad.png ../initial/2_toy2/3testo_nonad_subplots.png ../initial/2_toy2/3_ensemble_05_05.png ../initial/2_toy2/3scan.png ../initial/2_toy2/3scan_20.png ../initial/99_misc/2_lambdas.png; do cp $i plots_diskdisp; done
    \documentclass[twocolumn,twocolappendix]{aastex63}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        graphicx,
        ae, aecompl,
        booktabs,
    }
    \usepackage[T1]{fontenc}

@@

In this paper, we consider a wide range of initial planetary obliquities in the
Millholland-Batygin dissipating disk scenario, and examine how the obliquity
evolves toward the ``final'' value as the exterior disk dissipates. We provide
an analytical framework for understanding the final planetary obliquity for
arbitrary initial spin-disk misalignment angles. We also consider various
dissipation timescales, and examine both ``adiabatic'' (slow disk dissipation)
and ``non-adiabatic'' evolution. \textcolor{Corr}{Our chief new results include
that: (i) careful accounting of phase space area across separatrix crossings can
analytically describe the rich dynamics of adiabatic evolution, and (ii) the
concept of ``partial adiabatic resonance advection'' fully captures dynamics in
the non-adiabatic limit.} We calibrate our analytical results with numerical
calculations.

@@

We define the dimensionless time $\tau$ as
\begin{equation}
    \tau \equiv \alpha t,
\end{equation}
and the frequency ratio $\eta$
\begin{align}
    \eta \equiv{}& -\frac{g}{\alpha}\nonumber\\
        ={}& 2.08 \p{\frac{k_{\rm p}}{k_{\rm qp}}}
            \p{\frac{\rho_{\rm p}}{\mathrm{g/cm}^3}}
            \p{\frac{M_{\rm d}}{0.01 M_{\odot}}}
            \p{\frac{a_{\rm p}}{5\;\mathrm{AU}}}^{9/2}\nonumber\\
        &\times
            \p{\frac{r_{\rm d}}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{M_{\odot}}}^{-3/2}
            \p{\frac{P_p}{10\;\mathrm{hrs}}}
            \cos I ,\label{eq:eta}
\end{align}
% g = 1 / (51249 yr) 1 / (3 / 4 * (0.01) / (30 AU)^3 * sqrt(G * (solar mass) * (5AU)^3))
% alpha = 1/(106700 yr) 1 / (3/2 * (1 solar mass / (5AU)^3) / (4*pi / 3 * 1g/cm^3) * (2pi / (10 hours)))
% ((3 * 0.01 / 4) * (1 / 30AU)^3 * (10 hours * sqrt(G * (solar mass)) / (2pi)) / (1.5 * (1 solar mass / (5AU)^(9/2)) / (4 * pi / 3 * (1g/cm^3))))
where $\rho_p = 3M_p/(4\pi R_p^3)$ and $P_p = 2\pi/\Omega_p$ is the planet's
rotation period. In Eq.~\eqref{eq:eta}, we have introduced the fiducial values
of variable parameters for the application considered in this paper.
Eq.~\eqref{eq:dsdt_rot} then becomes
\begin{equation}
    \p{\rd{\uv{s}}{\tau}}_{\rm rot} = \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        - \eta\p{\uv{s} \times \uv{l}_{\rm d}}. \label{eq:dsdt_base}
\end{equation}

@@

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Trajectories in zone II librate about CS2 while those in
zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] \textcolor{Corr}{for $I = 5^\circ$}, where warmer colors
    denote more positive values. The black solid line is the separatrix, which
    only exists for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III),
    divided by the separatrix, are labeled. The Cassini states are denoted by
    filled circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The
    interior of the separatrix, shaded in grey, is formally only defined for
    $\eta < \eta_{\rm c}$, but we may identify the points in phase space that
    flow into zone II when evolved forward in time (decreasing $\eta$
    adiabatically); this is the shaded region in panel (a), bounded by the black
    dotted line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically. If we define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},
    \end{align}
\end{subequations}
\textcolor{Corr}{then the areas for $\eta < \eta_{\rm c}$ are given by
\citep{ward2004I}}
\begin{subequations}\label{se:area_ward}
    \begin{align}
        \mathcal{A}_{\rm I} &= 2\pi\p{1 - z_0} - \frac{\mathcal{A}_{\rm II}}{2},
            \label{eq:A1}\\
        \mathcal{A}_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan
            \frac{1}{\chi}, \label{eq:A2}\\
        \mathcal{A}_{\rm III} &= 2\pi\p{1 + z_0} - \frac{\mathcal{A}_{\rm
            II}}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$\mathcal{A}_{\rm j}\p{\eta > \eta_{\rm c}} = \mathcal{A}_{\rm j}(\eta_{\rm
c})$. The boundary between these extended zones is denoted by the dashed black
line in panel (a) of Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/1_areas.png}
    \caption{Plot of fractional areas of each of the zones
    $\mathcal{A}_{\rm j}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}
    \textcolor{Corr}{for $I = 5^\circ$}. The colored dotted lines correspond to
    small $\eta$ approximations used in Appendix~\ref{s:ad_approx}. The colored
    dashed lines for $\eta > \eta_{\rm c}$ are the effective values of
    $\mathcal{A}_{\rm II}, \mathcal{A}_{\rm III}$ for $\eta > \eta_{\rm c}$,
    denoting the points that would flow into either area under adiabatic
    decrease of $\eta$ from $\eta > \eta_{\rm c}$ (see the text). The vertical
    black dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}]
    and the values of $\eta$ for which $\mathcal{A}_{\rm II}$ is maximized
    ($\eta_{\rm \max, II}$) and for which $\mathcal{A}_{\rm III}$ is minimized
    ($\eta_{\rm \min, III}$, Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_05_35.png}
    \caption{Top: The final spin obliquity $\theta_{\rm f}$ as a function of the
    initial spin-disk misalignment angle $\theta_{\rm sd, i}$ for systems
    evolving from initial $\eta_{\rm i} \gg 1$ to $\eta_{\rm f} \ll 1$,
    \textcolor{Corr}{for $I = 5^\circ$}. The blue dots are results of numerical
    calculations (Section~\ref{ss:ad_ensemble}), and the colored tracks are
    semi-analytical results (Section~\ref{ss:zone_transitions}). Bottom: The
    probabilities of different outcomes. Where a particular $\theta_{\rm sd, i}$
    corresponds to multiple tracks, the system evolves probabilistically. The
    track that a particular system evolves along in a numerical simulation can
    be measured by examining its final obliquity. The dots represent the
    inferred probabilities from measured final obliquities in our simulations,
    while the colored tracks denote the semi-analytic probability of the system
    evolving along each track. There are five regimes of $\theta_{\rm sd, i}$
    values for which different tracks are accessible. In both plots, the
    vertical dashed black lines denote semi-analytical calculations of the
    boundaries of these regimes (see Section~\ref{ss:zone_transitions}), while
    the black dotted lines represent analytical approximations valid in the
    small-$\theta_{\rm sd, i}$ limit (see
    Appendix~\ref{s:ad_approx}).}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_20_35.png}
    \caption{Same as the top panel of Fig.~\ref{fig:ad_ensemble} but for $I
    = 20^\circ$ and with fewer annotations.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
\textcolor{Corr}{Note that $A$ can be negative when $\rdil{\phi}{t} < 0$, unlike
the unsigned areas $\mathcal{A}_{\rm i}$ [Eqs.~\eqref{se:area_ward}] which are
positive by definition.} This definition
of $A$ has two advantages: (i) it is continuous across transitions from
circulating to librating that cross the North pole ($\cos \theta = 1$), and (ii)
the areas of the three zones are equal in absolute value to the expressions
given in Eqs.~\eqref{se:area_ward}. The path over which the integral is taken is
either a libration or circulation cycle. When $\eta_{\rm i} \gg 1$, trajectories
librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm sd}$, meaning they
enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is studied \textcolor{Corr}{in the
adiabatic limit} by \citet{henrard1982} and \citet{henrard1987}. Their results
may be summarized as follows: if zone $i$ is shrinking while adjacent zones $j,
k$ are expanding such that the sum of their areas is constant, the probabilities
of transition from zone $i$ to zones $j$ and $k$ are given by\footnote{
\textcolor{Corr}{Interestingly, the probabilistic nature of separatrix crossing
is not because the outcome depends on the phase $\phi$ of the system at
crossing. Instead, the transition the system takes is determined by the value of
$H$ at the start of a separatrix-crossing orbit \citep{henrard1982}. Since $H$
only changes by $\propto \epsilon$ over a circulation/libration cycle, $\sim
\epsilon$ changes to initial conditions affect the final outcome. As such, as
$\epsilon \to 0$, the outcome effectively becomes probabilistic.}}
\begin{subequations}\label{eq:henrard_hop}
    \begin{align}
        \Pr\p{i \to j} = -\frac{\pdil{\mathcal{A}_{\rm
            j}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}, \\
        \Pr\p{i \to k} = -\frac{\pdil{\mathcal{A}_{\rm
            k}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}.
    \end{align}
\end{subequations}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$.
Eqs.~\eqref{eq:henrard_hop} can be used in conjunction with
Eqs.~\eqref{se:area_ward} to understand for what initial conditions each track
can be observed and with what probabilities.

@@

As $\epsilon$ increases (see Fig.~\ref{fig:3_ensemble_05_25}), nonadiabaticity
manifests as a larger scatter of final obliquities near the tracks predicted
from adiabatic evolution. This scatter first sets in for trajectories starting
in zone III, as these trajectories encounter the separatrix at larger $\eta$
compared to those originating in zone II\@. This means the obliquity of CS2
$\theta_2$ is smaller for these trajectories, and the adiabaticity criterion is
stricter [see Eq.~\eqref{eq:ad_constr}]. Physically, approaching the
adiabaticity criterion corresponds to the separatrix crossing process becoming
sensitive to the \emph{phase} of the libration/circulation cycle at the
crossing \textcolor{Corr}{(as noted in Section~3.2.2, the phase does not affect
separatrix crossing in the adiabatic limit)}: if the trajectory crosses the
separatrix when the obliquity is at its maximum, the final obliquity will also
be relatively larger.

As $\epsilon$ increases further (see Fig.~\ref{fig:3_ensemble_05_15}) but still
marginally satisfies the weak adiabaticity criterion [Eq.~\eqref{eq:ad_constr}],
the scatter in $\theta_{\rm f}$ continues to widen. The horizontal banded
structure of the final obliquities is a consequence of even stronger phase
sensitivity during separatrix crossing: trajectories cross the separatrix at
similar phases evolve to similar final obliquities that only depend weakly on
on $\theta_{\rm sd, i}$. \textcolor{Corr}{Finally, the bottom edge
of the data and the III $\to$ I track deviate very noticeably. This
non-adiabatic effect is the result of the separatrix evolving significantly
within the separatrix-crossing orbit, as the tracks computed in
Section~\ref{s:ad} assume $\eta$ is constant throughout the separatrix-crossing
orbit.}
